---
title: "Intro_to_package_building"
output: html_notebook
---
  
### Chapters  
  
1. The R Package Structure  
2. Documenting Packages  
3. Checking and Building R-packages  
4. Adding Unit Tests to R-packages  
  
### 1. The R Package Structure  
    
In this chapter, we will learn the basics of creating an R package. We will learn about the structure of R packages, set up a package, and write a function and include it in your package. We will also learn about the metadata stored in the DESCRIPTION and NAMESPACE files.  
  
**The structure of an R package**  
  
An R package can include:  
  
* Functions  
* Data  
* Documentation  
* Vignettes  
* Tests  
  
You don´t need all this.  
  
**Mandatory components**  
  
As a minimum our packages must include:  
  
* R directory - where you put the R code  
* man directory - where you store your documentation  
* NAMESPACE file - contains info about functions that the package imports from other packages and functions that the package makes available to its users  
* DESCRIPTION file - contains metadata about the package, author, purpose, current version, dependencies    
  
**Packages for easy creation of packages!**  
  
We dont need to do all this for hand, we will use packages that are designed for package building.  
  
* `devtools` , helps us stick to the right package structure  
* `roxygen2` , makes it simple to create docs in man directory and manage NAMESPACE file  
  
What we need to do our self is to write the functions we want and update the information in the template description file.  
  
**Setting up an R package**  
  
We will use the `devtools` package alot to maintain the structure of the package but also to write tests. The aim of `devtools` is to make development as easy as possible for package developers. Core devtools functions:  
  
* `create()`  
* `document()`  
* `check()`  
* `build()`  
* `test()`  
  
**create()**  
  
The `create()` function will create the basic structure, generating a NAMESPACE file, DESCRIPTION FILE and the R directory. It does not create the man directory yet, that happens later. To use `create()` you need to tell where to create the package and what name it should have. In below example we will create a packages called `simutils` in a new folder in the current work directory. The name of the folder we provide becomes the name of the package.

```{r}
# set work directory for where to create the new package, run in console!
```

Note: Avoid names already taken on **`CRAN`**    
```{r}
# Do not run
library(devtools)
create("simutils")
```
Running above `create("simutils")` will launch a new RStudio instance and create a new folder in the work directory:  
  
`/simutils` , a new folder  
  
In this folder **the following objects are** created:  
  
* /R , folder  
* .gitignore  
* .Rbuildignore  
* DESCRIPTION  
* NAMESPACE  
* simutils.Rproj  
  
Practice!  
  
**The structure of an R package**  
  
You can use `devtools` to create the basic structure of an R package by using the `create()` function.

There are some optional arguments to this function but the main one that you will use is the path argument. You use this to specify where your package will be created and the name that your package will take.

If you want to create the package in your current working directory, as you often will, you just need to supply the name for the package. When naming your package remember to think about:
  
* If the name is already taken by another package.  
* Whether the name makes it clear what the package does.  
  
Instructions:  
* Use the `create()` function to create a new R package called `datasummary`.  
* Use the `dir()` function to see what files and directories are created in your package.  
```{r}
# DO NOT RUN IF THE FOLDER datasummary ALREADY EXISTS!
library(devtools)

# Use the create function to set up your first package
create("datasummary")

# Take a look at the files and folders in your package
dir("datasummary")
```
Now let's take a look at the contents of your package in more detail
  
**Contents of an R package**  
  
When you create an R package there are a number of files and folders that it can contain. This can include folders to store data, user guides (known as vignettes), tests (more on these later in the course) and demos among others. As a minimum there are only two directories that must be included along with two additional files -- the `DESCRIPTION` and the `NAMESPACE`.

Of the two compulsory directories one contains all of the user guides for your functions. As you will see later in the course this is created in a special way for you. The other is included by the `create()` function. Which of the following is the second compulsory directory?  
  
Possible Answers:  
A. data  
B. tests  
C. inst  
D. R  
Correct: D  
That's right - all the code for your package goes in the R directory!  
    
**Writing a simple function**  
  
Whilst there are packages that contain only data, typically packages are created to collect together functions for performing a specific task.  
  
For our package you are going to keep the functions simple. You are going to create a package that produces custom summary output for your data.  
  
Instructions:  
* Write a function, `numeric_summary()`, that takes two arguments, a vector `x` and the logical `na.rm`.  
* Include a check to your function so it fails with an error message if the vector `x` is not numeric.  
* The function should return a `data.frame` containing the minimum, median, standard deviation and maximum values.  
* Check that your function works using the `Ozone` column of the `airquality` data and removing missing values.  
  
```{r}
# Create numeric_summary() function
numeric_summary <- function(x, na.rm = TRUE) {

    # Include an error if x is not numeric
    if(!is.numeric(x)){
        stop("Data must be numeric")
    }
    
    # Create data frame
    data.frame( min = min(x, na.rm = na.rm),
                median = median(x, na.rm = na.rm),
                sd = sd(x, na.rm = na.rm),
                max = max(x, na.rm = na.rm))
}

# Test numeric_summary() function
numeric_summary(airquality$Ozone, na.rm = TRUE)
```
Now let's look at how you include your function in a package.  
  
**Including functions in a package**  
  
Once you have written your function code you need to save it in the `R` directory of your package. Typically you can do that by saving an R script in the usual manner (i.e. "Save As").

In the instance that you already have objects created, as you did in the last exercise, that you want to write to the R directory you can do this programmatically. You can use the function `dump()` to send a named R function to a particular file. The two arguments that you need to pass to this function are the name of the R object, as a character string, and the path to the file that you want to create, including the extension `.R`.  
  
Instructions:  
* Use the `dir()` function again to see what is currently available in the `R` directory of `datasummary`.  
* Using the `dump()` function, write the `numeric_summary()` function to the R directory, saving the file as `numeric_summary.R`.  
* Print the contents of the `R` directory again.  
```{r}
# What is in the R directory before adding a function?
dir("datasummary/R")

# Use the dump() function to write the numeric_summary function
dump("numeric_summary", file = "datasummary/R/numeric_summary.R")

# Verify that the file is in the correct directory
dir("datasummary/R")
```
The function has now been saved correctly!  
  
**DESCRIPTION and NAMESPACE files**  
  
DESCRIPTION file  

* General package information  
* Author and maintainer details  
* Package dependencies  
* License  
  
Example in folder datasummary/DESCRIPTION and another one below.
```{r}
Package: datasummary
Title: What the Package Does (One Line, Title Case)
Version: 0.0.0.9000
Authors@R: C( 
    person("John", "Doe", 
           email = "first.last@domain.com", role = c("aut", "cre")),
    person("Tom", "Hill",
           email = "first.last@domain.com", role = c("aut", "ctn")))
Description: Solves some analytical problems blazing fast.
Depends:
    R (>= 3.4.2)
License: GPL-2
Encoding: UTF-8
LazyData: true
Roxygen: list(markdown = TRUE)
RoxygenNote: 7.1.0
Imports: dplyr, 
    purrr
Suggests: testhat,
    knitr,
    rmarkdown
VignetteBuilder: knitr
```
  
**NAMESPACE file**  
  
The NAMESPACE file controls two important things:  
  
* Functions or packages imported for use by our package to be able to run  
* Functions exported by our package for people to use  
  
We do NOT edit this file by hand!  
We will manage it with all our other documentation.  
  
Example NAMESPACE file in folder datasummary/DESCRIPTION and another one below.
```{r}
# Generated by roxygen2: do not edit by hand

export(na_counter)
export(sample_from_data)
import(dplyr)
import(purrr)
```
We are exporting na_counter to our users and we need dplyr, purrr so it works.
  
A package should not have the same name as an existing package and its name must only contain letters, numbers, or dots.  
  
The DESCRIPTION file contains details about the other packages your package uses, it doesn't include which functions are in your package.  
  
From the role = c('cph','ctb','cre','aut') the maintainer is the role 'cre'.  
  
**Optional directories**  
  
We have seen what the mandatory directories are that we need but we can also include many other things like:  
  
* Data  
* Vignettes  
* Tests  
* Compiled code  
* Translations  
  
**Data**  
  
Most of this can be included with the `use_data()` function. Example with Data.
```{r}
# DO NOT RUN
sim_data <- data.frame(
  ID = 1:10,
  Value = sample(1:11, 10),
  Apples = sample(c(TRUE,FALSE), 10, replace = TRUE)
)
```
Data is stored in the R directory as and R data file. We can call the `use_data()` function to add the data from an R object. The function needs to know the name of the R object and the location of the package.
```{r}
# DO NOT RUN
use_data(sim_data, pkg = "simutiles")
```
  
**Vignette**  
  
If we want to create a vignette, which is like a user-guid we can call the `use_vignette` function.
```{r}
use_vignette("my_first_vignette", pkg = "simutils")
```
This will create a directory called vignette and create a template vignette for us to edit with the name we provide. `devtools` will create an `rmarkdown` template in which `devtools` will create a special header which we should leave as is so that R identifies our vignette.  
  
**Best practice for structuring code**  
  
You cannot create just any structure or your package wont build!  
  
Guidelines for the R directory:  
* No subdirectories  
* Don't have everything in one script  
 * For small packages you might want to have one file for each function, where each file name reflects the function.  
* Don't have a large number of small files.  
* As package gets larger, group similar functions together into one file.  
* A good practice is to have large functions on their own while have similar, smaller functions grouped together.  
  
Practise!  
  
**The use_* functions**  
  
Beyond the required structure you can include a number of additional directories containing elements such as vignettes (user guides), data and unit tests. The `devtools` package makes it really simple for you to add to the package structure by providing a series of `use_*` functions. For example, `use_data()` and `use_vignette()`. Note that when adding vignettes, it's best not to include any spaces in the vignette name.

When you are adding data you need to provide the name of the data object along with the argument `pkg`, giving the path to the package that you want to put your data in.  
  
Instrucitons:  
* Using the `dir()` function, check what directories are currently in the package?  
* The `weather` dataset has been created for you. Add it to your package.  
* Add a vignette titled `"Generating_Summaries_with_Data_Summary"`.
```{r}
# OMA. Create weather data
weather = data.frame(Day = as.integer(c(1,2,3,4,5,6,7)),
                     Temp = as.integer(c(22,19,18,24,15,21,14)),
                     Weather = as.character(c("Snow","Snow","Sun","Rain",
                                              "Rain","Snow","Rain")))
library(devtools)
#In console set run: setwd("C:/repo/dc/datasummary")

#What is in the package at the moment?
dir("datasummary")

# Add the weather data
use_data(weather, pkg = "datasummary", overwrite = TRUE)

# What directories do you now have in your package now?
dir("datasummary")

```
This error is due to this project should be in a root folder, not as sub-folder to dc.  
  
**Best practice for structuring code**  
  
A typical R package contains a number of functions that you need to maintain. Whilst there are no strict rules around how you should structure code in a package you generally want to avoid having all of your code in a single script. As you can't have sub-directories you also need to think carefully about how you name the file so that you can find your code again in the future.

Suppose you were to write another function for your package that takes all numeric columns in your data and returns a data frame of all of their summary statistics. What would be the best way to structure this code?  
  
Instructions:  
* Send the new `data_summary()` function to a new file taking the same name as the function (You should use the dump() function here)
```{r}
data_summary <- function(x, na.rm = TRUE){
  
  num_data <- dplyr::select_if(x, .predicate = is.numeric)
  
  purrr::map_df(num_data, .f = numeric_summary, na.rm = TRUE, .id = "ID")
  # numeric_summary function was defined and dumped previously to our package
  
}

# Test function
data_summary(weather)

# Write the function to the R directory
dump("data_summary", file = "datasummary/R/data_summary.R")
```
Now that you've learned about the R package structure, the next step is to see how to document your package.

End of chapter 1.  
  
  
### 2. Documenting Packages  
  
**Introduction to roxygen2**  
  
The documentation of a package is very important for the user of the packages.  
  
* Help files  
  
Function help files list all of the arguments to the function and detail the purpose and use of each and other information. To create these files we can use a package called `roxygen2`. It allows us to ad a header to a function containing all the important information about the function. When you then run an appropriate build function the documentation in the form of help files is built automatically from these roxygen headers. Let´s look at one example:  
  
**roxygen headers**  
  
```{r}
#' Sample from data
#'
#' Sample rows form a dataset.
#'
#' This function has been designed to sample from the rows of a two dimensional
#' data set, returning all columns of sampled rows. Sampiing is done with 
#' replacement by default
#'
#' @param data A data frame or matrix from which rows are to be sampled
#' @param size Numeric. Number of rows to return
#' @param replace Logical. Sample with replacement? TRUE by default.
#'
#' @authoer Firstname Lastname
#'
#' @import dplyr
#'
#' @returns A data set of the same type as input with \code{size} rows.
#' @export  
#'
#' @examples
#' sample_from_data(airquality, size=10)
sample_from_data <- function(data, size, replace=TRUE) {
  
  if(!is.numeric(size)){
    stop("size must be numeric value")
  }
  
  if(is.matrix(data)){
    data = as.data.frame(data)
  }
}
```
Each row in the roxygen header start with a hash and a single quote `#'`.  
Following this you can indicate particular components of the helpfile.  
Let´s look at some components in details.   
  
* The first paragraph of the header is treated as title of the help file `#' Sample from data`, this should be short and no longer than one sentence.  
* The second paragraph of the header is treated as a description for the help page `#' Sample rows form a dataset.` and should include a brief description of the function  
* The third paragraph `#' This function has been designed..`, is treated as the detailed section. We might want to provide more info about complex parameters of functionality here.  
* For everything else, we need to use special tags preceeded by the `@` symbol.  
* The `@param` tags define each of the function arguments and the corresponding help text that the user will see in the help documentation. It is usefull to specify the object type and give a short description: `@param size Numeric. Number of rows to return`  
* Using other functions in packages requires us to import the package, not just loading it using the `library()` function. The `@import` tag is used to specify which package must be imported, e.g: `@import dplyr`. When you import a funciton you make it´s NAMESPACE and functions available to your package. In this example the `dplyr` function is imported so that the functions of the latter package can be use in the body of our package.  
  
Practise!  
  
**A simple function header**  
  
The `roxygen` headers are included in the same script as the function code. You use roxygen comments `#'` to identify a line as part of the roxygen header. You can include regular R comments in the header if you wish, using the usual comment character, `#`.

The first three lines of the header have special meaning and you don't need to use tags to identify them. The first three lines are:  
  
* title  
* description  
* details  
  
Since you don't mention any tags to identify the first three lines, it is necessary that you seperate each one of them by a new line. For example:
```{r}
#' Title goes here
#'
#' Description goes here
#'
#' Details go here
```
Instructions:  
  
* Add the title "Numeric Summaries" to your header.  
* Add the following short description of the function: "Summarises numeric data and returns a data frame containing the minimum value, median, standard deviation, and maximum value."  
```{r}
# Add a title and description
#' Numeric Summaries
#'
#' Summarises numeric data and returns a data frame containing the minimum value, median, standard deviation, and maximum value
numeric_summary <- function(x, na.rm){

    if(!is.numeric(x)){
        stop("Data must be numeric")
    }
    
    data.frame( min = min(x, na.rm = na.rm),
                median = median(x, na.rm = na.rm),
                sd = sd(x, na.rm = na.rm),
                max = max(x, na.rm = na.rm))
}
```
Now let's look at how to document function arguments.  
  
**Documenting function arguments**  
  
One of the most important aspects of a function to document are the arguments.  
  
With roxygen you do this with the `@param` tag. You follow the tag with the argument name and then any details the user needs to know.  
  
If there is a lot of information that you need to provide about an argument you might want to consider putting this in the details section instead.  
  
Normally you must document every argument of your function, but noe just document the first argument and come back to the second one later!  
  
Instructions:  
  
* Add an appropriate tag to document the first argument of the ´numeric_summary()` funciton.  
* Add the following details to this tag: "a numeric vector containing the values to summarize".  
```{r}
#' Numeric Summaries
#'
#' Summarises numeric data and returns a data frame containing the minimum value, median, standard deviation, and maximum value.
#'
# Add appropriate tag and details to document the first argument
#'@param x a numeric vector containing the values to summarize.

numeric_summary <- function(x, na.rm){

    if(!is.numeric(x)){
        stop("data must be numeric")
    }
    
    data.frame( min = min(x, na.rm = na.rm),
                median = median(x, na.rm = na.rm),
                sd = sd(x, na.rm = na.rm),
                max = max(x, na.rm = na.rm))
}
```
It's important to document your parameters so others (or your future self!) can see how to use your functions.  
  
**Importing other packages**  
  
If you wish to use functions from other packages, you should import them in your `roxygen` header. You can use `@import` to import an entire package, or `@importFrom` to import a single function. To show you how `@importFrom` works, we have imported the `gather()` function from the `tidyr` package for you (although it's not needed in this case).  
  
Instructions:  
  
* Update the `data_summary()` function to import packages purrr and dplyr.
```{r}
#' Summary of Numeric Columns
#'
#' Generate specific summaries of numeric columns in a data frame
#' 
#' @param x A data frame. Non-numeric columns will be removed
#' @param na.rm A logical indicating whether missing values should be removed

#' @import purrr
#' @import dplyr 
#' @importFrom tidyr gather
data_summary <- function(x, na.rm = TRUE){
  
  num_data <- select_if(x, .predicate = is.numeric) 
  
  map_df(num_data, .f = numeric_summary, na.rm = na.rm, .id = "ID")
  
}
```
Now that you've looked at importing functions, let's now take a look at exporting them.  
  
**How to export functions**  
  
Video.  
*Exported functions* (can be loaded by the end-user)  
* visible to the end user
* key package functionality  
  
*Non-exported functions* (functions we dont want end-user using directly)  
* not visible to end user
* utility functions  
  
When documenting using `roxygen` the exported functions are automatically listed in the NAMESPACE file.  
  
Examples of exported and non-exported functions.  
  
Below is no `@export` tag so it is a utility function and not exported, is used by other functions in the package but not a function we want end-user to access:
```{r}
#' Cound NAs in a vector
#' 
#' @param x A vector
#' 
#' @return Number of NAs in x
#' 
#' @examples
#'  sumNa(airquality$Ozone)
sum_na <- function(x) {
  sum(is.na(x))
}
```

Another example is exported, it has an `@export` tag, and it is available for the end-user:
```{r}
#' Cound NAs in a data set
#' 
#' @param data A data frame or matrix
#' 
#' @return Number of NAs in x
#'
#' @import purr
#'
#' @return Vector of NA counts
#' @export
#'
#' @examples
#' na_counter(airquality)
na_counter <- function(data) {
  
  stopifnot(is.matrix(data) | is.data.frame(data))
  
  if(is.matrix(data)) {
    data = as.data.frame(data)
  }
  purrr::map_int(data. sum_na)
  }
```
Now let´s look at how we refer to these functions.  
Imaging the `simutils` packages has been built and loaded.
```{r}
# Do not run
library(simutils)
na_counter(airquality) # this would work, since it is exported

sum_na(airquality$Ozone) # this would give an error, cannot be found, not exported!
```
Only the exported at loaded!

Calling non-exported functions is possible. Need to use `package_name:::` three colons.
```{r}
# Do not run
simutils:::sum_na(airquality$Ozone) # This should work since using three colons :::
```
  
How to use roxygen headers to export functions.  
```{r}
#'
#' @export # Use this tag to export it, if we dont want to export it then remove this tag
#'
```
Export best practice  
  
Exporting a function makes it available for the end users of your package. Functions that are not exported are not directly available to your end users. It's a really useful way to hide low level functions that you write to break up the main functionality. We want to export high level functions for calculating all summaries of supplied data in our example above. Not utility functions.  
  
**Exporting functions**  
  
Although the NAMESPACE file is used to identify functions that are exported you use roxygen tags to create this file. This makes it really easy to see, right next to the function, if your users will be able to work with it or not.  
  
Instructions:  
* Add an export tag to the `data_summary()` function.
```{r}
#' Summary of Numeric Columns
#'
#' Generate specific summaries of numeric columns in a data frame
#' 
#' @param x A data frame. Non-numeric columns will be removed
#' @param na.rm A logical indicating whether missing values should be removed
#' @import dplyr
#' @import purrr
#' @importFrom tidyr gather
#' @export
data_summary <- function(x, na.rm = TRUE){
  
  num_data <- select_if(x, .predicate = is.numeric) 
  
  map_df(num_data, .f = numeric_summary, na.rm = na.rm, .id = "ID")
  
}
```
Only functions we want the end user to have direct access to should be exported. Now let's look at the other things we can document.  
  
**Documenting other elements**  
  
Video.  
Documenting examples, we can do this to demonstrate how to use our function. We should provide at least one example for any function that is exported and for any functions that have complicated arguments. An example should not produce an error and there are some requirements on CRAN about examples. Check these out before submitting. This is how to add an example in the roxygen header:  
```{r}
#'
#' @examples 
#' sample_from_data(airqualiity, size=10)
```
Example are automatically run during the package checking stage (talk about this later). In some cases we want to add an example but without the example being run. You can do this:
```{r}
#'
#' @examples
#' \dontrun{
#' sample_from_data(airqualiity, size=10)
#' }
```
We can use the `\dontrun` markup tag for this. This is applicable e.g if the function relies on an API-key or being connected to a database. 
  
We an use the `@return` tag to describe to the user what to expect from the user if it is data, graphics of any other output:
```{r}
#'
#' @param size Numeric. Number of rows to return.
#'
#' @return A data set of the same type as input with \code{size} rows.
#'
#' @examples
#' sample_from_data(airqualiity, size=10)
#' }
```
We used a sentence to desribe returned output and the markup tag `\code{size}` to indicate that we are using one of the parameters.  
  
Additional documentation.  
```{r}
#'
#' @author John Doe
#'
```
`@authoer` tag to write who wrote the package.  
`@notes` tag to add some info if the funciton is experimental or likely to change.  
  
**Adding examples**  
  
Adding examples
Examples are the way that your end users will learn how to work with your code. You should provide at least one example for any function that is exported. If there are any tricky arguments to your function you might also want to consider examples for those.

Examples should ideally be easily reproducible and should not cause an error. If you are going to publish your package to CRAN then examples should also run in a reasonable time, CRAN sets this to around 15 minutes but it's worth double checking the recommendations at the time of submission.

You add examples using the `@examples` tag. Examples will typically span multiple lines. This is fine as `roxygen` will assume that everything is part of the examples until it finds another tag.  
  
Instructions:  
* Add an examples tag to the `data_summary()` function header.
* Add an example of running the function using the `iris` data.
* Add a second example of running the function with the `airquality` data, not removing missing values.
```{r}
#' Data Summary for Numeric Columns
#'
#' Custom summaries of numeric data in a provided data frame
#'
#' @param x A data.frame containing at least one numeric column
#' @param na.rm A logical indicating whether missing values should be removed
#' @import dplyr
#' @import purrr
#' @importFrom tidyr gather
#' @export
#'
#' @examples 
#' data_summary(iris)
#' data_summary(airquality, na.rm=FALSE)
data_summary <- function(x, na.rm = TRUE){
  
  num_data <- select_if(x, .predicate = is.numeric) 
  
  map_df(num_data, .f = numeric_summary, na.rm = na.rm, .id = "ID")
  
}
```
Examples can really help users understand your function.  
  
**Documenting function return values**  
  
You document the return value of a function using the tag `@return`. This is where you can tell users what they can expect to get from the function, be that data, a graphic or any other output.

When you write documentation you may want to format the text to look like code, link to other functions or, particularly for return objects, format as a bulleted list. You can do this using special formatting:  
  
* For code you use `\code{text to format}`
* To link to other functions you use `\link[packageName]{functionName}`, although note the package name is only required if the function is not in your package
* To include an unordered list you use `\itemize{}`. Inside the brackets you mark new items with `\item` followed by the item text.  
  
Instructions:  
* Document the return value with the following description: "This function returns a data.frame including columns:". Format the word `data.frame` as code.
* Add 5 bullets for each of the 5 column names in the order they are returned (the function is loaded if you want to run it).
```{r}
#' Data Summary for Numeric Columns
#'
#' Custom summaries of numeric data in a provided data frame
#'
#' @param x A data.frame containing at least one numeric column
#' @param na.rm A logical indicating whether missing values should be removed
#' @import dplyr
#' @import purrr
#' @importFrom tidyr gather
#' @export
#' @examples
#' data_summary(iris)
#' data_summary(airquality, na.rm = FALSE)
#'
## Update the details for the return value
#' @return This function returns a \code{data.frame} including columns:
#' \itemize{
#'  \item ID
#'  \item min
#'  \item median
#'  \item sd
#'  \item sd
#'  \item max
#' }
#'
#' @export
data_summary <- function(x, na.rm = TRUE){
  
  num_data <- dplyr::select_if(x, .predicate = is.numeric) 
  
  purrr::map_df(num_data, .f = numeric_summary, na.rm = na.rm, .id = "ID")
  
}

data_summary(iris)
```
Next we're going to look at additional documentation.  
  
**Additional documentation**  
  
Beyond the functions arguments and return values there are many additional items that can be provided in function help files. From who wrote that function to other functions that you might be interested in using. We will also see later how we can identify dependencies using the `roxygen` header.  
  
Other tags that we can use include:  
  
* `@author` to identify who wrote the function.  
* `@seealso` to list other functions that may be of interest to users.
* `@notes` to add any other notes relating to the function (e.g. if its experimental, likely to change etc.)  
  
Instructions:  
* Add in the author of the `data_summary()` function as `"My Name <myemail@example.com>"`.
* List the `summary()` function (from the `base` package) after adding the author.
```{r}
#' Summary of Numeric Columns
#' Generate specific summaries of numeric columns in a data frame
#'
#' @param x A data frame. Non-numeric columns will be removed
#' @param na.rm A logical indicating whether missing values should be removed
#' @import dplyr
#' @import purrr
#' @importFrom tidyr gather
#' @export
#' @examples
#' data_summary(iris)
#' data_summary(airquality, na.rm = FALSE)
#' 
#' @return This function returns a \code{data.frame} including columns: 
#' \itemize{
#'  \item ID
#'  \item min
#'  \item median
#'  \item sd
#'  \item max
#' }
#'
## Add in the author of the `data_summary()` function
#' @author "My Name <myemail@example.com>"
## List the `summary()` function (from the `base` package)
#' @seealso \link[base]{summary}
data_summary <- function(x, na.rm = TRUE){
  
  num_data <- dplyr::select_if(x, .predicate = is.numeric) 
  
  purrr::map_df(num_data, .f = numeric_summary, na.rm = na.rm, .id = "ID")
  
}
```
You don't have to document the return value, although it can be useful to the user to do so. Now that you know how to document the functions in a package, let's look at how to document the package itself.  
  
**Documenting a package**  
  
Video. 
We have documented functions using `roxygen` headers, now we will look at how to document other components of our package. We can also generate the package level documentaiton using `roxygen2`.  
  
*Package level documentation*  
  
We need to add tag `@docType` and `@name`. The header it self needs to be followed by `"_PACKAGE"` key word because all headers need to be followed by some `r`code. It is good practice to create a separate r-script which takes the package name that contains this header. This is how it might look in the simutils package:
```{r}
#' simutils: A package for performing common simulation task
#'
#' This package provides functionality for a variet of simulation tasks,
#' and plotting tools for viewing the results.
#'
#' @author John Doe \email{john.doe@blablabla.com}
#' @docType package
#' @name simutils
"_PACKAGE"

```  
We save this file as a `simutils.R` file in the `R` directory.
  
**Minimum level of documentation**  
  
For each function, document:  
* Title
* Description
* Arguments
* Exported (for exported functions only)  
  
You can also document other things that are going to be usefull to the end user of our package.  
  
*Documenting data objects*  
Data that we want any user to see can be included in the data directory. `use_data` function in `devtools` will help us compress the data and to handle the structure of the package properly. Below we take a data frame `sim_dat` and compress it and put it into the data directory all with just a single function call.
```{r}
# DO NOT RUN
use_data(sim_dat, pkg = "simutils")
```
  
Need to have some documentation for our added data sets. We can also do this using `roxygen2` using `@format`and `@source`. Here is an example.
```{r}
#' sim_data data set
#'
#' We made some data for the package
#'
#' @format A data.frame with 3 columns
#' \describe{
#' \item{ID}{ID value}
#' \item{Value}{Measured value in pounds}
#' \item{Apples}{Logical. Do they like apples}
#' }
#' @source Simulated Data
#'
"sim_dat"

```  
  
*Creating man files*  
  
Once we have created function headers and the package help files we can create the `.rd` files for the package, which go into the `man` directory, using the `document()` function in `devtools`.
```{r}
# DO NOT RUN
document("simutils")
```
This will create or updating all of the `.rd` files as well as the `NAMESPACE`file.  
  
**Adding package documentation**  
  
In addition to the individual functions you can also document the whole package.  
  
When you document a package you can use all of the same tags as for functions but the problem that you have is that there is no R code to document. Instead you put the keyword "_PACKAGE" underneath your package header.  
  
You should also use the @docType and @name tags to indicate that this is package level documentation, and specify your package name.

You should then save the package documentation in the R directory in a file ending with .R with the same name as the package.  
  
Instrucitons:  
* Write a package header for your `datasummary` package with the title "Custom Data Summaries".
* Include the description "Easily generate custom data frame summaries".
* Add the relevant tags required for package level documentation.
* Ensure that the `"_PACKAGE"` keyword is added in the appropriate location.
```{r}
#' datasummary: Custom Data Summaries
#'
#' Easily generate custom data frame summaries
#'
#' @docType package
#' @name datasummary
"_PACKAGE"
```
  
**Documenting data objects**  
  
If you include a data set in your package you must also provide documentation for it. As with package level documentation, you should include a title and description.  
  
With a data set, once again, there is no function object to document. You instead give the name of the dataset as a character string, for instance `"airquality"`. Additionally there are two tags that you need to use:

* `@format` to describe the format of the data. This is often used with the describe format.
* `@source` to identify where the data came from.  
You should then save the package documentation in the R directory with a suitable name (in the above case, either `airquality.R` or `data.R`).  
  
The `weather` dataset is available in your workspace. Run `print(weather)` to view it.
  
Instructions:  
* Update the `@format` tag to include the correct number of rows and columns in `weather`.
* Add descriptions of all columns in the `weather` dataset that you added to your package in the last chapter.
* Add a `@source` tag to and describe the data as "Randomly generated data".
```{r}
#' Random Weather Data
#'
#' A dataset containing randomly generated weather data.
#'
#' @format A data frame of 7 rows and 3 columns
#' \describe{
#'  \item{Day}{Numeric values giving day of the week, 1 = Monday, 7 = Sunday}
#'  \item{Temp}{Numeric values giving temperature in degrees Celsius}
#'  \item{Weather}{Character values describing the weather on that day}
#' }
#' @source Randomly generated data
"weather"
```
Now you've created your documentation, the next step is to build it.  
  
**Creating man files**  
  
Once you've created your documentation using `roxygen` headers, you need to build your documentation. You can use the `document()` function from the `devtools` package to generate your documentation, supplying the path to the package as the first argument. The generated documentation will appear in the `man` directory. Once you've documented your package, you can view the help files just like you can with any of your existing packages.    
Instructions:  
* Generate documentation for your package.
* Look at the contents of the `man` directory.
* View the documentation for the `data_summary()` function.
* View the documentation for the `weather` dataset.
```{r}
# OMA: Will generate error, since have not dumped the data into the data folder, this should contain weather.rda, and might be missing something else in some other folder.

# Also in dc this was printed: We are overriding the actual `help()` function in this exercise so you can view the documentation. You don't have to do this on your local system.


# Generate package documentation
document("datasummary")

# Examine the contents of the man directory
dir("datasummary/man")

# View the documentation for the data_summary function
help("data_summary")

# View the documentation for the weather dataset
help("weather")
```
Now your package is fully documented, it's time to run some final checks before building it.  
  
  
### 3. Checking and Building R-packages  
  
**Why check an R package?**  
  
Video.

    
JATKA TÄSTÄ  
https://campus.datacamp.com/courses/developing-r-packages/checking-and-building-r-packages?ex=1
